A comprehensive approach to fitting multivariate latent Markov models and latent transition models: An implementation and applications

### 2008-11-03 papers/jss/hmmlta.tex
added papers directory (lots of fighting with svn ...)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@234 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[a4paper,12pt,doc]{apaIngmar} % nobf, doc, apa, man

\usepackage[]{amsmath, amsfonts, amstext, amsthm}

\usepackage{amssymb}

\usepackage[]{graphics}

\usepackage{graphicx}

\usepackage{epsfig}

\usepackage{epstopdf}

\newcommand{\citep}{\cite}

\newcommand{\citet}{\citeA}

\newcommand{\mat}{\mathbf}

\newcommand{\vc}{\mathbf}

\title{DMM4: Dependent Mixture Models in S4 Classes}

\author{Ingmar Visser}
\date{\today}

\affiliation{Department of Psychology, University of Amsterdam\\

 Correspondence concerning this article should be addressed to:
 \\
 Ingmar Visser \\
 Department of Psychology, University of Amsterdam \\
 Roetersstraat 15 \\
 1018 WB Amsterdam \\
 phone: +31 (20) 5256735 \\
 fax: +31 (20) 6390279 \\
 email: i.visser@uva.nl}

\renewcommand{\baselinestretch}{1.5}

\shorttitle{DMM4: Dependent Mixture Models in S4}

\abstract{DMM4 implements mixture models and latent Markov models for
mixed indicator from various distributions.  Optimization can be
subjected to general linear constraints and is done using nlm, optim
or NPSOL. Covariates can be regressed on model parameters through
identity or logit link functions.}

@@

\bibliographystyle{apacite}

@@

\newpage

@@

The model can be seen as an extension of several other models, in
particular, the hidden Markov model (HMM), the latent class model
(LCM), the latent transition model (LTM) and also the simple Markov
model (MM).  Markov models (and their subsequent refinements) have a
long history in psychology, mainly in the psychology of learning
\citep{Wickens1982}.  These models have in common that they are all
used to describe one or more sequences of responses $O_{1}, \ldots,
O_{T}$, where each of the $O_{t}$'s may be multivariate.  In this
paper mixed categorical and continuous observations are considered.

The model concerned in this paper is the combination of the latent
class model and the hidden Markov model.  There is quite some work
already on this model, but it is usually limited to small numbers of
repeated meausurements, in which case this model is usually called the
latent transition model: it is the repeated measurements extension of
the latent class model.  It is customary in those models to have
time-varying (transition) parameters.  In the time series literature
on the other hand, where hidden Markov models are frequently used,
transition parameters are taken to be time-homogeneous.  This
assumption is necessary when analyzing single time-series since at
each time $t$ only a single transition is observed.  The multi-variate
latent Markov model is defined by the following elements:
\begin{enumerate}
	\item a set $\vc{S}$ of latent states $S_{i},\, i=1, \ldots , n$,

	\item a matrix $\mat{A}$ of transition probabilities $a_{ij}$ for
	the transition from state $S_{i}$ to state $S_{j}$,

	\item a set $\vc{B}$ of observation functions $b_{j}(\cdot)$ that
	provide the conditional probabilities associated with latent state
	$S_{j}$,

	\item a vector $\pmb{\pi}$ of latent state initial probabilities
	$\pi_{i}$.
\end{enumerate}
When transitions are added to the latent class model, it is more
appropriate to refer to the classes as states.  The word class is
rather more associated with a stable trait-like attribute whereas a
state can change over time.

@@

For computing the likelihood and its partial derivatives, I follow the
strategy used by \cite{Lystig2002}.  They developed a method to
compute exact gradients and information matrix by reformulating the
forward procedure of the Baum-Welch algorithm for estimating hidden
Markov model parameters.  Scores and information are computed at a
single pass through the data, using the intermediate variables from
the forward recursion.  Their article only discusses univariate time
series.  Our aim in this section is to generalize their results to
multiple dependent and independent time series.  With the gradients
available, it is possible to use general purpose optimization routines
that allow box constraints and arbitrary (non-) linear constraints.

@@

The loglikelihood of hidden Markov models is usually computed by the
so-called forward-backward algorithm \citep{Baum1966,Rabiner1989}, or
rather by the forward part of this algorithm.  \cite{Lystig2002}
change the forward algorithm in such a way as to allow computing the
gradients of the loglikelihood at the same time.  They start by
rewriting the likelihood as follows (for ease of exposition the
dependence on the model parameters is dropped):
\begin{equation}
	L_{T} = Pr(\vc{O}_{1}, \ldots, \vc{O}_{T}) = \prod_{t=1}^{T}
	Pr(\vc{O}_{t}|\vc{O}_{1}, \ldots, \vc{O}_{t-1}),
	\label{condLike}
\end{equation}
where $Pr(\vc{O}_{1}|\vc{O}_{0}):=Pr(\vc{O}_{1})$.  Note that for a
simple Markov chain these probabilities reduce to
$Pr(\vc{O}_{t}|\vc{O}_{1},\ldots, \vc{O}_{t-1}) =
Pr(\vc{O}_{t}|\vc{O}_{t-1})$.  The log-likelihood can now be expressed
as:
\begin{equation}
	l_{T} = \sum_{t=1}^{T} \log[Pr(\vc{O}_{t}|\vc{O}_{1}, \ldots,
	\vc{O}_{t-1})].
	\label{eq:condLogl}
\end{equation}

To compute the log-likelihood, \cite{Lystig2002} define the following 
(forward) recursion:
\begin{align}
	\phi_{1}(j) &:= Pr(\vc{O}_{1}, S_{1}=j) = \pi_{j}
	b_{j}(\vc{O}_{1}) \label{eq:fwd1} \\
\begin{split}
	\phi_{t}(j) &:= Pr(\vc{O}_{t}, S_{t}=j|\vc{O}_{1}, \ldots,
	\vc{O}_{t-1}) \\
	&= \sum_{i=1}^{N} [\phi_{t-1}(i)a_{ij}b_{j}(\vc{O}_{t})] \times
	(\Phi_{t-1})^{-1},
	\label{eq:fwdt} 
\end{split} 
\end{align}
where $\Phi_{t}=\sum_{i=1}^{N} \phi_{t}(i)$.  Combining
$\Phi_{t}=Pr(\vc{O}_{t}|\vc{O}_{1}, \ldots, \vc{O}_{t-1})$, and
equation~(\ref{eq:condLogl}) gives the following expression for the
log-likelihood:
\begin{equation}
	l_{T} = \sum_{t=1}^{T} \log \Phi_{t}.
\end{equation}

@@

% %Using the equations above for the likelihood of single and mixture 
% models, arriving at the posterior probabilities of membership of the 
% components can be easily accomplished.

% %The posterior probability is the probability that data $\vc{O}$ are 
% generated by mixture component $k$, i.e., $Pr(C=k|\vc{O})$, where $C$ 
% denotes the mixture component. This probability can be expressed as:
%\begin{eqnarray}
%	Pr(C=k|\vc{O}) &=& \frac{Pr(C=k) Pr(\vc{O}|C=k)}
%	{\sum_{k=1}^{K} Pr(C=k)Pr(\vc{O}|C=k)} \\
%	&=& \frac {p_{k} l_{TMC=k}}
%	{\sum}
%\end{eqnarray}

@@

See equations 10--12 in \cite{Lystig2002} for the score recursion
functions of the hidden Markov model for a univariate time series.
Here the corresponding score recursion for the multivariate mixture
case are provided.  The $t=1$ components of this score recursion are
defined as (for an arbitrary parameter $\lambda_{1}$):
\begin{align}
\psi_{1}(j_{k};\lambda_{1}) &:=  \fpp Pr(\vc{O}_{1}|S_{1}=j_{k}) \\
\begin{split} 
	&= \left[  \fpp p_{k} \right] \pi_{j_{k}}b_{j_{k}}(\vc{O}_{1}) + 
	p_{k}\left[ \fpp \pi_{j_{k}} \right] b_{j_{k}}(\vc{O}_{1}) \\
	& \qquad  + p_{k}\pi_{j_{k}} \left[ \fpp 
b_{j_{k}}(\vc{O}_{1})\right],
\end{split} \label{eq:psi1}
\end{align}
and for $t>1$ the definition is:
\begin{align}
	\psi_{t}(j_{k};\lambda_{1}) & = \frac{\fpp Pr(\vc{O}_{1}, \ldots,
	\vc{O}_{t}, S_{t}=j_{k})} {Pr(\vc{O}_{1}, \ldots, \vc{O}_{t-1})}
	\\
	\begin{split} 
		& = \sum_{i=1}^{n_{k}} \Bigg\{
		\psi_{t-1}(i;\lambda_{1})a_{ij_{k}} b_{j_{k}}(\vc{O}_{t}) \\
			&\qquad +\phi_{t-1}(i) \left[ \fpp a_{ij_{k}} \right]
			b_{j_{k}} (\vc{O}_{t}) \\
			&\qquad +\phi_{t-1}(i)a_{ij_{k}}  \left[ \fpp b_{j_{k}} 
			(\vc{O}_{t}) \right] \Bigg\} 
			\times (\Phi_{t-1})^{-1}.
		\end{split} 
	\label{eq:psit}
\end{align}

Writing this out in matrix notation (and dropping the $k$-subscripts 
for components) we get for $t=1$ and for each $k$:
\begin{eqnarray}
	\pmb{\psi}_{1}(\lambda_{1}) = \vc{\fpp} p_{k} \times \pmb{\pi} 
	\otimes \vc{b}(O_{1})+ p_{k} \times \vc{\fpp}\pmb{\pi} \otimes
	\vc{b}(O_{1}) + p_{k} \times \pmb{\pi} \otimes \fpp \vc{b}(O_{1}),
\end{eqnarray}
and for $\pmb{\psi}_{t}$ we get: 
\begin{equation}
	\begin{split} 
		\pmb{\psi}_{t}(\lambda_{1}) &= \left[ \mat{A}^{t} \times
		\pmb{\psi_{t-1}}(\lambda_{1}) \right ] \otimes \vc{b}(O_{t}) +
		\left [ \left[ \fpp \mat{A} \right ]^{t} \times
		\pmb{\phi}_{t-1} \right ] \otimes \vc{b}(O_{t}) \\
		& \qquad + \left [ \mat{A}^{t} \times \pmb{\phi}_{t-1} \right]
		\otimes \fpp \vc{b}(O_{t}).
	\end{split}
\end{equation}

Using above equations, \cite{Lystig2002} derive the following equation 
for the partial derivative of the likelihood:
\begin{equation}
	\fpp l_{T}=
	\frac{\mathbf{\Psi}_{T}(\lambda_{1})}{\mathbf{\Phi}_{T}},
\end{equation}
where $\Psi_{t}=\sum_{k=1}^{K} \sum_{i=1}^{n_{k}} 
\psi_{t}(j_{k};\lambda_{1})$. 
Starting from the equation from the logarithm of the likelihood, this 
is easily seen to be correct: 
\begin{eqnarray}
	\fpp \log Pr(\vc{O}_{1}, \ldots, \vc{O}_{T}) &=& Pr(\vc{O}_{1},
	\ldots, \vc{O}_{T})^{-1} \fpp Pr(\vc{O}_{1}, \ldots, \vc{O}_{T})
	\\
	&=& Pr(\vc{O}_{1}, \ldots, \vc{O}_{T})^{-1} \Psi_{T}(\lambda_{1})
	Pr(\vc{O}_{1}, \ldots, \vc{O}_{T-1}) \\
	&=& \frac{Pr(\vc{O}_{1}, \ldots, \vc{O}_{T-1})}{Pr(\vc{O}_{1},
	\ldots, \vc{O}_{T})} \Psi_{T} (\lambda_{1}) \\
	&=&  \frac{\mathbf{\Psi}_{T}(\lambda_{1})}{\mathbf{\Phi}_{T}}.
\end{eqnarray}

@@

For an $c$-category item $l$ with response probabilities $p_{1}, 
\ldots, p_{c}$ we have the following:
\begin{align} 
	\fpp b_{j_{k}}(O_{t}^{l})= \left\{ 
	\begin{array}{ll} 
		1 & \mbox{if $\lambda_{1} = p_{i}$ \& $O_{t}=i$}  \\ 
		0 & \mbox{otherwise} 
	\end{array} 
	\right\}
\end{align}

%\newpage
%\section*{Author note}
%We would like to thank many people.

\bibliography{all}

>>>>

\documentclass[a4paper,man,nobf]{apa}
 \usepackage[]{amsmath, amsfonts, amstext, amsthm}
 \usepackage[oztex, final]{graphics}
 \usepackage[]{epsfig}

\usepackage[round]{natbib}

\title{A comprehensive approach to fitting multivariate latent Markov 
models and latent transition models: An implementation and 
applications}

\author{Ingmar Visser}
\affiliation{Developmental Processes Research Group, Department of 
Psychology, University of Amsterdam}
\date{\today}

% \acknowledgements{Author notes etc.}   
 \rightheader{Depedent Mixtures}                    % defaults to title
\shorttitle{Depedent Mixtures}                      % defaults to title
% \leftheader{Authors}

%\batchmode

%\newcommand{\citep}{\cite}
%\newcommand{\citet}{\citeA}

\newcommand{\vc}{\mathbf}
\newcommand{\mat}{\mathbf}

\abstract{In this article, an outline is presented for a program that 
computes parameter estimates with standard errors for the 
multivariate hidden Markov model. The model itself can be seen as an 
extension and generalization of both the hidden or latent Markov 
model and latent class models. The parameter procedure is different 
from the traditional approach with these models. This approach allows 
modeling of arbitrarily long time series efficiently and without 
exponentially growing computational effort. Moreover, because 
analytical gradients and Hessian are available, models can easily be 
fitted subject to arbitrary linear and non-linear constraints. 
Multi-group and mixture analyses are also possible. A number of 
illustrative examples are presented: 1) mixture analysis of 
(Markovian) learning data , 2) analysis of sleep stages from EEG, and 
3) intra- and inter-individual differences in personality structure.}

%\keywords{latent Markov model; latent class model; latent transition model; mixtures; gradients; 
%%%% Hessian; estimation}

@@

%\bibliographystyle{plainnat}

@@

@@

The model can be seen as an extension of several other models, in 
particular, the hidden Markov model (HMM), the latent class model 
(LCM), the latent transition model (LTM) and also the simple Markov 
model (MM). Markov models (and their subsequent refinements) have a 
long history in psychology, mainly in the psychology of learning 
\citep{Wic82}. These models have in common that they are all used to 
describe one or more sequences of responses $O_{1}, \ldots, O_{T}$, 
where each of the $O_{t}$'s may be multivariate. In this paper mixed 
categorical and continuous observations are considered.

The model concerned in this paper is the combination of the latent 
class model and the hidden Markov model. There is quite some work 
already on this model, but it is usually limited to small numbers of 
repeated meausurements, in which case this model is usually called 
the latent transition model: it is the repeated measurements 
extension of the latent class model. It is customary in those models 
to have time-varying (transition) parameters. In the time series 
literature on the other hand, where hidden Markov models are 
frequently used, transition parameters are taken to be 
time-homogeneous.  This assumption is necessary when analyzing single 
time-series since at each time $t$ only a single transition is 
observed. The multi-variate latent Markov model is defined by the 
following elements:
\begin{enumerate}
  	\item a set $\vc{S}$ of latent states $S_{i},\, i=1, \ldots , n$,
  	\item a matrix $\mat{A}$ of transition probabilities $a_{ij}$ for 
the transition 
	from state $S_{i}$ to state $S_{j}$,
  	\item a set $\vc{B}$ of observation functions $b_{j}(\cdot)$ that 
provide the conditional 	probabilities associated with latent state 
$S_{j}$, 
  	\item a vector $\pmb{\pi}$ of latent state initial probabilities 
$\pi_{i}$.
\end{enumerate}
When transitions are added to the latent class model, it is more 
appropriate to refer to the classes as states. The word class is 
rather more associated with a stable trait-like attribute whereas a 
state can change over time.

@@

For computing the likelihood and its partial derivatives, I follow 
the strategy used by \cite{Lys2002}. They developed a method to 
compute exact gradients and information matrix by reformulating the 
forward procedure of the Baum-Welch  algorithm for estimating hidden 
Markov model parameters. Scores and information are computed at a 
single pass through the data, using the intermediate variables from 
the forward recursion. Their article only discusses univariate time 
series. Our aim in this section is to generalize their results to 
multiple dependent and independent time series. With the gradients 
available, it is possible to use general purpose optimization 
routines that allow box constraints and arbitrary (non-) linear 
constraints.

@@

The loglikelihood of hidden Markov models is usually computed by the 
so-called forward-backward algorithm \citep{Bau66,Rab89}, or rather 
by the forward part of this algorithm.
\cite{Lys2002} change the forward algorithm in such a way as to allow 
computing the gradients of the loglikelihood at the same time. They 
start by rewriting the likelihood as follows (for ease of exposition 
the dependence on the model parameters is dropped): 
\begin{equation}
	L_{T} = Pr(\vc{O}_{1}, \ldots, \vc{O}_{T}) = \prod_{t=1}^{T} 
Pr(\vc{O}_{t}|\vc{O}_{1}, 
	\ldots, \vc{O}_{t-1}), 
	\label{condLike}
\end{equation}
where $Pr(\vc{O}_{1}|\vc{O}_{0}):=Pr(\vc{O}_{1})$. Note that for a 
simple Markov chain these probabilities reduce to 
$Pr(\vc{O}_{t}|\vc{O}_{1},\ldots, 
\vc{O}_{t-1})=Pr(\vc{O}_{t}|\vc{O}_{t-1})$.
The log-likelihood can now be expressed as:
\begin{equation}
	l_{T} = \sum_{t=1}^{T} \log[Pr(\vc{O}_{t}|\vc{O}_{1}, \ldots, 
\vc{O}_{t-1})].
	\label{eq:condLogl}
\end{equation}

To compute the log-likelihood, \cite{Lys2002} define the following 
(forward) recursion:
\begin{align}
	\phi_{1}(j) &:= Pr(\vc{O}_{1}, S_{1}=j) = \pi_{j} b_{j}(\vc{O}_{1}) 
	\label{eq:fwd1} \\
\begin{split}
	\phi_{t}(j) &:= Pr(\vc{O}_{t}, S_{t}=j|\vc{O}_{1}, \ldots, 
\vc{O}_{t-1}) \\
	&= \sum_{i=1}^{N} [\phi_{t-1}(i)a_{ij}b_{j}(\vc{O}_{t})] \times 
(\Phi_{t-1})^{-1},
	\label{eq:fwdt} 
\end{split} 
\end{align}
where $\Phi_{t}=\sum_{i=1}^{N} \phi_{t}(i)$. Combining 
$\Phi_{t}=Pr(\vc{O}_{t}|\vc{O}_{1}, \ldots, \vc{O}_{t-1})$, and 
equation~(\ref{eq:condLogl}) gives the following expression for the 
log-likelihood:
\begin{equation}
	l_{T} = \sum_{t=1}^{T} \log \Phi_{t}.
\end{equation}

@@

%Using the equations above for the likelihood of single and mixture 
models, arriving at the posterior probabilities of membership of the 
components can be easily accomplished.

%Nou, nee dus!

%The posterior probability is the probability that data $\vc{O}$ are 
generated by mixture component $k$, i.e., $Pr(C=k|\vc{O})$, where $C$ 
denotes the mixture component. This probability can be expressed as:
%\begin{eqnarray}
%	Pr(C=k|\vc{O}) &=& \frac{Pr(C=k) Pr(\vc{O}|C=k)}
%	{\sum_{k=1}^{K} Pr(C=k)Pr(\vc{O}|C=k)} \\
%	&=& \frac {p_{k} l_{TMC=k}}
%	{\sum}
%\end{eqnarray}

@@

See equations 10--12  in \cite{Lys2002} for the score recursion 
functions of the hidden Markov model for a univariate time series. 
Here the corresponding score recursion for the multivariate mixture 
case are provided. The $t=1$ components of this score recursion are 
defined as (for an arbitrary parameter $\lambda_{1}$):
\begin{align}
\psi_{1}(j_{k};\lambda_{1}) &:=  \fpp Pr(\vc{O}_{1}|S_{1}=j_{k}) \\
\begin{split} 
	&= \left[  \fpp p_{k} \right] \pi_{j_{k}}b_{j_{k}}(\vc{O}_{1}) + 
	p_{k}\left[ \fpp \pi_{j_{k}} \right] b_{j_{k}}(\vc{O}_{1}) \\
	& \qquad  + p_{k}\pi_{j_{k}} \left[ \fpp 
b_{j_{k}}(\vc{O}_{1})\right],
\end{split} \label{eq:psi1}
\end{align}
and for $t>1$ the definition is:
\begin{align}
\psi_{t}(j_{k};\lambda_{1})  & =  \frac{\fpp Pr(\vc{O}_{1}, \ldots, 
\vc{O}_{t}, S_{t}=j_{k})}
			{Pr(\vc{O}_{1}, \ldots, \vc{O}_{t-1})}  \\
\begin{split} 
			 & =  %\sum_{k=1}^{K} klopt dit?? is \fpp a_{ij} wel altijd 0 voor andere componenten??
			 \sum_{i=1}^{n_{k}} \Bigg\{ \psi_{t-1}(i;\lambda_{1})a_{ij_{k}} 
			 b_{j_{k}}(\vc{O}_{t}) \\ 
			 &\qquad +\phi_{t-1}(i) \left[ \fpp a_{ij_{k}} \right] b_{j_{k}} 
(\vc{O}_{t}) \\
			&\qquad +\phi_{t-1}(i)a_{ij_{k}}  \left[ \fpp b_{j_{k}} 
(\vc{O}_{t}) \right] \Bigg\} 
			\times (\Phi_{t-1})^{-1}.
\end{split} \label{eq:psit}
\end{align}

Writing this out in matrix notation (and dropping the $k$-subscripts 
for components) we get for $t=1$ and for each $k$:
\begin{eqnarray}
	\pmb{\psi}_{1}(\lambda_{1}) = \vc{\fpp} p_{k} \times \pmb{\pi} 
\otimes \vc{b}(O_{1})+
		p_{k} \times \vc{\fpp}\pmb{\pi} \otimes \vc{b}(O_{1}) +
		p_{k} \times \pmb{\pi} \otimes \fpp \vc{b}(O_{1}),
\end{eqnarray}
and for $\pmb{\psi}_{t}$ we get: 
\begin{equation}
\begin{split} 
	\pmb{\psi}_{t}(\lambda_{1}) 
	&= \left[ \mat{A}^{t} \times \pmb{\psi_{t-1}}(\lambda_{1})  \right ] 
\otimes \vc{b}(O_{t}) +
	\left [ \left[ \fpp \mat{A} \right ]^{t} \times \pmb{\phi}_{t-1} 
\right ] \otimes \vc{b}(O_{t}) \\
	& \qquad + \left [ \mat{A}^{t} \times \pmb{\phi}_{t-1}  \right ] 
\otimes \fpp \vc{b}(O_{t}).
\end{split} 
\end{equation}

Using above equations, \cite{Lys2002} derive the following equation 
for the partial derivative of the likelihood:
\begin{equation}
	\fpp l_{T}= 	
		\frac{\mathbf{\Psi}_{T}(\lambda_{1})}{\mathbf{\Phi}_{T}},
\end{equation}
where $\Psi_{t}=\sum_{k=1}^{K} \sum_{i=1}^{n_{k}} 
\psi_{t}(j_{k};\lambda_{1})$. 
Starting from the equation from the logarithm of the likelihood, this 
is easily seen to be correct: 
\begin{eqnarray}
	\fpp \log Pr(\vc{O}_{1}, \ldots, \vc{O}_{T}) &=& Pr(\vc{O}_{1}, 
\ldots, \vc{O}_{T})^{-1} 
	\fpp Pr(\vc{O}_{1}, \ldots, \vc{O}_{T}) \\
	&=& Pr(\vc{O}_{1}, \ldots, \vc{O}_{T})^{-1} \Psi_{T}(\lambda_{1}) 
Pr(\vc{O}_{1}, \ldots, \vc{O}_{T-1}) \\
	&=&  \frac{Pr(\vc{O}_{1}, \ldots, \vc{O}_{T-1})}{Pr(\vc{O}_{1}, 
\ldots, \vc{O}_{T})}  \Psi_{T} (\lambda_{1}) \\
	&=&  \frac{\mathbf{\Psi}_{T}(\lambda_{1})}{\mathbf{\Phi}_{T}}.
\end{eqnarray}

@@

For an $c$-category item $l$ with response probabilities $p_{1}, 
\ldots, p_{c}$ we have the following:
\begin{align} 
        \fpp b_{j_{k}}(O_{t}^{l})= \left\{ 
	\begin{array}{ll} 
		1 & \mbox{if $\lambda_{1} = p_{i}$ \& $O_{t}=i$}  \\ 
		0 & \mbox{otherwise} 
            \end{array} \right. 
\end{align}

\subsection{Observed information}

\newcommand{\fpt}{\frac{\partial}{\partial \lambda_{2}} }

\newcommand{\fps}{\frac{\partial^{2}}{\partial\lambda_{1}\partial\lambda_{2}}}

To compute the observed information, again the formulae by 
\citet{Lys2002} are applied and generalized to the m-variate mixture 
case. Define the information recursion as follows:
\begin{align}
\begin{split}
	\omega_{1}(j_{k};\lambda_{1},\lambda_{2}) 
	=  & \, \fps Pr(\vc{O}_{1}, S_{1}=j_{k}) = \fps p_{k} \pi_{j_{k}} 
b_{j_{k}}(\vc{O}_{1})
	\\  = & \, \fpp \left[ \fpt p_{k} \pi_{j_{k}} b_{j_{k}}(\vc{O}_{1}) 
\right] 
	\\  =  & \fpp  \bigg\{  \left[ \fpt p_{k} \right]  \pi_{j_{k}} 
b_{j_{k}}(\vc{O}_{1}) +
		\, p_{k} \left[ \fpt \pi_{j_{k}} \right]  b_{j_{k}}(\vc{O}_{1}) + 
	\\  &  \, p_{k} \pi_{j_{k}} \left[ \fpt  b_{j_{k}}(\vc{O}_{1}) 
\right] \bigg\}
	\\  =  & \,  \left[ \fps p_{k} \right] \pi_{j_{k}} 
b_{j_{k}}(\vc{O}_{1}) + 
		\left[ \fpt p_{k} \right] \cdot \left[\fpp \pi_{j_{k}} \right] 
b_{j_{k}}(\vc{O}_{1}) +
	\\  &   \, \left[ \fpt p_{k} \right] \pi_{k} \left[ \fpp b_{j_{k}} 
(\vc{O}_{1}) \right] +
		\left[ \fpp p_{k} \right] \pi_{j_{k}} b_{j_{k}}(\vc{O}_{1}) + 
	\\ &   \, p_{k} \left[ \fps \pi_{j_{k}} \right] 
b_{j_{k}}(\vc{O}_{1}) +
		p_{k} \left[ \fpt \pi_{j_{k}} \right] \cdot \left[ \fpp 
b_{j_{k}}(\vc{O}_{1}) \right] +
	\\ &   \, \left[ \fpp p_{k} \right] \pi_{j_{k}} \left[ \fpt 
b_{j_{k}}(\vc{O}_{1}) \right] +
		p_{k} \left[ \fpp \pi_{j_{k}} \right] \cdot \left[ \fpt 
b_{j_{k}}(\vc{O}_{1}) \right] +
	\\ &   \, p_{k} \pi_{j_{k}} \left[ \fps b_{j_{k}}(\vc{O}_{1}) 
\right].
\end{split}
\end{align}

Similarly, for $t>1$, the information recursion is defined as:
\begin{align}
\begin{split}
	\omega_{t}(j_{k};\lambda_{1},\lambda_{2}) =  
	&\frac{\fps Pr(\vc{O}_{1}, \ldots, 
\vc{O}_{t},S_{t}=j_{k})}{Pr(\vc{O}_{1}, \ldots, \vc{O}_{t-1})}
	\\ = &\sum_{k=1}^{K} \sum_{i=1}^{n_{k}} \Bigg\{ 
	\omega_{t-1}(i;\lambda_{1},\lambda_{2}) a_{ij_{k}} 
b_{j_{k}}(\vc{O}_{t}) +
	\psi_{t-1}(i;\lambda_{1}) \left[ \fpt a_{ij_{k}} \right] 
b_{j_{k}}(\vc{O}_{t}) +
	\\  &\psi_{t-1}(i;\lambda_{1})  a_{ij_{k}}  \left[ \fpt 
b_{j_{k}}(\vc{O}_{t}) \right] +
	\psi_{t-1}(i;\lambda_{2}) \left[ \fpp a_{ij_{k}} \right] 
b_{j_{k}}(\vc{O}_{t}) + 
	\\ & \psi_{t-1}(i;\lambda_{2}) a_{ij_{k}} \left[ \fpp 
b_{j_{k}}(\vc{O}_{t})\right]  + 
	\phi_{t-1}(i) \left[ \fps a_{ij_{k}} \right]  b_{j_{k}}(\vc{O}_{t}) +
	\\ & \phi_{t-1}(i) a_{ij_{k}} \left[ \fps b_{j_{k}}(\vc{O}_{t}) 
\right]  +
	\phi_{t-1}(i) \left[ \fpp a_{ij_{k}} \right] \left[ \fpt 
b_{j_{k}}(\vc{O}_{t}) \right]  +
	\\ & \phi_{t-1}(i) \left[ \fpt a_{ij_{k}}\right] \left[ \fpp 
b_{j_{k}}(\vc{O}_{t}) \right]  \Bigg\} 
	\times \frac{1}{\Phi_{t-1}}
\end{split}
\end{align}

It is important here to provide the second order partial derivatives 
of the terms that are involved in the multi-variate case. These are 
the $b_{j_{k}}(\cdot)$ terms. Remember that these terms can be 
factored and hence the derivatives can be derived from repeatedly 
applying the chain rule for products and using equation 
\ref{partialProd} from above for the first order derivatives: 
\begin{align}
\begin{split}
	\fps b_{j_{k}} (\vc{O}_{t}) & = \fps \prod_{i=1}^{m} b_{j_{k}}^{i} 
(O_{t}^{i}) 
	\\ & = \fpp \left[ \fpt \prod_{i=1}^{m} b_{j_{k}}^{i} (O_{t}^{i})  
\right] 
	\\ & = \fpp \Bigg\{ \sum_{l=1}^{m} \left[ \prod_{i=1, \ldots, 
\hat{l}, \ldots, m} b_{j}(O_{t}^{i}) 
	\right] \times \fpt  [b_{j}(O_{t}^{l})] \Bigg\}  
	\\ & = \sum_{l=1}^{m} \left[ \fpp \prod_{i=1, \ldots, \hat{l}, 
\ldots, m} b_{j}(O_{t}^{i}) \right] 
	\times \fpt  \left[ b_{j}(O_{t}^{l}) \right] \; + 
	\\ &\phantom{=}\;\,  \sum_{l=1}^{m}  \left[ \prod_{i=1, \ldots, 
\hat{l}, \ldots, m} b_{j}(O_{t}^{i}) 
	\right] \times \fpp \fpt  \left[ b_{j}(O_{t}^{l})\right]  
	\\ & = \sum_{k=1,k\neq	l}^{m} \sum_{l=1}^{m}  \left[ \prod_{i=1, 
\ldots, \hat{l}, \hat{k}, \ldots, m}
	b_{j}(O_{t}^{i}) \right] \fpt  \left[ b_{j}(O_{t}^{k})\right]  \fpt  
\left[ b_{j}(O_{t}^{l})\right] \;+ 
	\\ &\phantom{=}\;\, \sum_{l=1}^{m}  \left[ \prod_{i=1, \ldots, 
\hat{l}, \ldots, m} b_{j}(O_{t}^{i}) 
	\right] \times \fps  \left[b_{j}(O_{t}^{l}) \right] 
\end{split}
\end{align}

In order to compute the above terms we need for arbitrary parameters 
$\lambda_{1}$ the terms $\fps  \lambda_{1}$. Fortunately, these are 
easy to compute as most of them are zero. In particular, with 
$\lambda_{1} \in \{\mat{A},\vc{p},\pmb{\pi} \}$, the double 
derivatives are zero and we only need them when 
$\lambda_{1}\in\{\mat{B}\}$.
These latter ones are again zero in the case of multinomial items and 
so only the gaussian observations remain which are given here in four 
cases:

%%p-k cases
\paragraph{Case 1: $\fps b_{j}(O_{t}^{l}), \lambda_{1}=\mu, 
\lambda_{2}=\mu$}

\begin{align} 
        \fps b_{j_{k}}(O_{t}^{l})=b_{j}(O_{t}^{l}) \left[ \left[ 
\frac{O_{t}^{l}-\mu}{\sigma^{2}} \right]^{2} - \frac{1}{\sigma^{2}} 
\right]
\end{align}

\paragraph{Case 2: $\fps b_{j}(O_{t}^{l}), \lambda_{1}=\mu, 
\lambda_{2}=\sigma$}

\begin{align} 
        \fps b_{j_{k}}(O_{t}^{l})=b_{j}(O_{t}^{l}) \left[ \left[ 
\frac{(O_{t}^{l}-\mu)^{2}}{\sigma^{3}} -
	\frac{1}{\sigma}\right]  \cdot \left[ 
\frac{O_{t}^{l}-\mu}{\sigma^{2}} \right] + 
	\left[ \frac{2\mu-2O_{t}^{l}}{\sigma^{3}} \right] \right]
\end{align}

\paragraph{Case 3: $\fps b_{j}(O_{t}^{l}), \lambda_{1}=\sigma, 
\lambda_{2}=\mu$}

\begin{align} 
        \fps b_{j_{k}}(O_{t}^{l})=b_{j}(O_{t}^{l})  \left[ \left[ 
\frac{(O_{t}^{l}-\mu)^{2}}{\sigma^{3}} -
	\frac{1}{\sigma}\right]  \cdot \left[ 
\frac{O_{t}^{l}-\mu}{\sigma^{2}} \right] + 
	\left[ \frac{2\mu-2O_{t}^{l}}{\sigma^{3}} \right] \right]
\end{align}

\paragraph{Case 4: $\fps b_{j}(O_{t}^{l}), \lambda_{1}=\sigma, 
\lambda_{2}=\sigma$}

\begin{align} 
        \fps b_{j_{k}}(O_{t}^{l})=b_{j}(O_{t}^{l}) \left[ \left[ 
\frac{(O_{t}^{l}-\mu)^{2}}{\sigma^{3}} -
	\frac{1}{\sigma}\right]^{2} + \left[  \frac{1}{\sigma^{2}} - 
\frac{3(O_{t}^{l}-\mu)^{2}}{\sigma^{4}}  \right]  \right]
\end{align}

\bibliography{psy,ingmar,hmm}

% \section*{Author note}
